Global patterns and drivers of vertebrate verticality

### 2024-08-21 verticality_future_mkd.Rmd
revision

---
author: "Lydia Soifer, David Edwards, Brett Scheffers"
title: "Global patterns and drivers of vertebrate verticality"
bibliography: "`r rbbt::bbt_write_bib('docs/r-refs.bib', overwrite = TRUE)`"
output: 
  officedown::rdocx_document:
    reference_docx: "style-guide.docx"
    mapstyles:
      Normal: ['First Paragraph']
plots:
  caption:
    style: Image Caption
    pre: 'Figure '
    sep: ': '
---

@@

Vertical stacking of species is an emblematic signature of biologically rich regions and is associated with patterns of species richness from local to global scales [@macarthurPopulationEcologyWarblers1958; @macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019; @bashamVerticalStratificationPatterns2023a]. At local scales, species assemblages may be stratified across vertical resource and climate gradients from beneath the ground to the top of the canopy [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023]. Tall, multi-layered forests with high foliage height diversity increase both the volume of habitat and the niche opportunities within the habitat, acting effectively as multi-floored apartments that enable species to vertically partition resources [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]. By reducing competition, vertical stratification may thus act as a mechanistic link to positive relationships between canopy height and species richness [@culbertInfluenceVerticalHorizontal2013; @fengForestCanopyHeight2020; @gillespieDistributionBirdSpecies2001; @wagnerRemotelySensedTree2023]. *concept fig* This hypothesis has been supported by positive relationships between verticality, defined here as the average vertical position of species in an assemblage, and species richness [@oliveiraVerticalStratificationInfluences2019]. However, canopy height and other metrics of vegetation structure do not consistently impact species richness at regional to global scales, with notable declines in the strength of the relationship at higher latitudes where the diversity of arboreal species generally declines [@larueDiversityVolumeRelationships2023; @rollLinkingVertebrateSpecies2015; @oliveiraVerticalStratificationInfluences2019]. Examining the impacts of climatic and vegetative drivers of vertical stratification may thus add nuance to our understanding of the impact of canopy height on species richness.

At regional to global scales, climate may constrain vertical stratification in some regions despite the presence of tall forests. Vertical climate gradients are characterized by fossorial and terrestrial habitats that exhibit lower thermal variability and greater moisture availability than arboreal habitats, which are exposed to hotter maximum temperatures, colder minimum temperatures, and lower humidity [@scheffersTropicalMountainPasses2018; @leahyVerticalNicheElevation2021; @defrenneGlobalBufferingTemperatures2019]. In high latitude forests, low minimum temperatures in the canopy may exceed physiological thresholds and constrain species to lower vertical positions [@oliveiraVerticalStratificationInfluences2019; @bakenSalamanderArborealityLimited2021]. Similarly, exposure to hot and dry conditions in the canopies of tropical lowland forests may reduce the suitability of arboreal habitats for heat or desiccation sensitive species, such as frogs [@scheffersIncreasingArborealityAltitude2013; @bashamVerticalStratificationCollapses2020]. Release of these climatic constraints at higher elevations in the tropics, where temperature declines and humidity increases, allows a greater number of species to occupy higher positions in the canopy [@scheffersIncreasingArborealityAltitude2013]. We therefore expect the impact of canopy height on verticality to vary across large spatial extents, exhibiting a positive impact where climate supports arboreal life strategies and a neutral impact where climatic conditions pose greater physiological challenges.

Spatial variation may also occur in the relationship between species richness and  verticality [@oliveiraVerticalStratificationInfluences2019]. Though prior studies have shown positive relationships between verticality and richness, the strength of this relationship is variable and generally more positive in tropical regions [@cabralFunctionalTraitsPhylogeny2022; @oliveiraVerticalStratificationInfluences2019]. Where climatic conditions allow canopy height to positively impact verticality, as may be expected in the wet tropics, the positive relationship between verticality and richness may be stronger because the addition of a vegetation layer allows the addition of arboreal species to the assemblage. In contrast, where taller canopies do not facilitate higher verticality, the relationship between verticality and richness may be weak (*concept fig*).

As the climate changes, constraints on arboreal life strategies may shift with potential impacts on assemblage verticality and its association with species richness. In tropical lowlands, increasing temperatures and declining dry season precipitation (*cite*) may pose physiological challenges for arboreal animals and cause them to disperse toward higher elevations, lower vertical positions, or become locally extirpated [@scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019]. These scenarios would reduce assemblage verticality in the lowlands, reflecting climate-induced shifts in assemblage thermal tolerance [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024]. At high latitudes, increasing minimum temperatures could reduce climatic constraints on arboreal species and facilitate range shifts that increase verticality. While changes in assemblage verticality have been observed across spatial environmental gradients and in response to seasonal climatic fluctuations [@scheffersIncreasingArborealityAltitude2013; @bashamVerticalStratificationCollapses2020], potential impacts of climate change on verticality have not been examined at a global scale.

@@

Despite widespread agreement that verticality influences biodiversity patterns, global patterns of verticality, its environmental drivers, and its relationship with species richness have not been comprehensively described for vertebrates. To improve our understanding of the nuances of the relationships between climate, vegetation structure, verticality, and species richness, we integrate range maps of `r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians with climate and vegetation data to examine global patterns and environmental drivers of vertebrate verticality in wooded biomes. We then examine whether the relationship between verticality and canopy height mediates the relationship between verticality and species richness. Finally, we predict how climate change may impact global patterns of verticality.

@@

We mapped the verticality and richness of vertebrate assemblages in wooded habitats using 50 km grid cells and a cylindrical equal area projection. Grid cells with fewer than five species were excluded. We identified wooded habitats by ecoregions including forests, woodlands, taiga, várzea, Yungas, savanna, thicket, and mallee as defined by Dinnerstein et al. [-@dinersteinEcoregionBasedApproachProtecting2017]. Range maps were obtained from the IUCN red list for mammals and amphibians [@iucnIUCNRedList2023], BirdLife International for birds [@birdlifeinternationalandhandbookofthebirdsoftheworldBirdSpeciesDistribution2022], and the Global Assessment of Reptile Distributions (GARD) for reptiles [@rollGlobalDistributionTetrapods2017; @caetanoAutomatedAssessmentReveals2022]. For birds, we used breeding plus resident ranges, because seasonality impacts the verticality of assemblages. To calculate the verticality of each assemblage (i.e., grid cell), we assigned each species a verticality score between zero and one and took the average verticality of all species whose ranges overlapped with the grid cell. For mammals, reptiles, and amphibians, verticality scores were based on their use of fossorial (0), terrestrial (0.5), and arboreal (1) habitats as defined in the TetrapodTraits database [@mouraPhylogenyinformedCharacterisationGlobal2024]. This dataset uses phylogenetic imputation to identify traits if trait data is not available in the literature. For species where imputation was necessary, we considered a species to use a given vertical stratum if the imputation probability exceeded 0.7. For birds, we calculated verticality scores based on their use of different foraging strata as identified in the EltonTraits database [@wilmanEltonTraitsSpecieslevelForaging2014]. We classified foraging below the water surface, around the water surface, and on the ground as 0.5, in the understory as 0.667, mid-high canopy as 0.8337, and in or above the upper canopy as 1. If species used multiple vertical habitats, vertical scores were averaged. For example, if a species occurred in terrestrial (0.5) and arboreal (1) habitats, it would receive a verticality score of 0.75. Species that occupied only aerial vertical niches or that did not occur in terrestrial habitats were excluded from the analysis.

Verticality exhibited a strong positive correlation with species richness. To examine spatial patterns of verticality independently of species richness, we calculated the standard effect size (SES) of mean verticality (\$ SES = (x_i - \bar{x})/sd(x)\$, where x is a vector of expected mean verticality scores given random selection of species from a species pool). We defined species pools as species within all wooded habitats in each biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017] and generated a null distribution of expected verticality given random chance for each grid cell using 100 simulations.

@@

We obtained present and future global climate data from CHELSA v2.1 for mean daily maximum air temperature of the warmest month, mean daily minimum air temperature of the coldest month, temperature seasonality, precipitation of the wettest month, precipitation of the driest month, and precipitation seasonality [@kargerDataClimatologiesHigh2021; @kargerClimatologiesHighResolution2017]. Present climate data represents the time period 1981-2010 and future climate data represents an ensemble of climate scenarios for SSP585 for the time period 2071-2100. Vegetation data included canopy height [@langHighresolutionCanopyHeight2023], vegetation density [@crowtherMappingTreeDensity2015], and vegetation complexity (the product of canopy height and vegetation density). Additionally, we included historical climate velocity, which represents the average rate at which thermal isotherms have shifted between the last glacial maximum and present climate conditions [@sandelInfluenceLateQuaternary2011] and has been shown to influence global patterns of verticality [@oliveiraVerticalStratificationInfluences2019]. All environmental data was resampled to the resolution and projection of the assemblage data.

@@

where $y_s$ is mean or SES verticality at location $s$ defined by x and y coordinates, $𝜇_s$ is the expected mean or SES verticality, and $𝑓^{-1}$ represents an inverse link function. We modeled SES verticality as a normal distribution and mean verticality as a beta distribution with a logit link function because values are constrained between 0 and 1. $𝜲^{main}_s$ is a vector of main effect covariates and 𝛽 represents a corresponding vector of estimated main effect coefficients. $α_g$ represents an optional random intercept by group g. $⍵_s$ represents a spatial random field at location $s$, and describes additional spatial variation in the data not explained by the environmental covariates or random intercept. $𝜲^{svc}_sζ_s$ represents an optional spatially varying coefficient, where $ 𝜲^{svc}$ represents the vector of spatially varying parameter values and $ζ_s$ represents the spatially varying coefficients, which are described using a random field such that the coefficients follow a multivariate normal distribution with mean zero and covariance ($𝚺_ζ$). The spatial range for the covariance function is shared with the range for the spatial random field. Both the spatial random field and the spatially varying coefficients are modeled as Gaussian random fields with the terms following a multivariate normal distribution with a covariance matrix modeled by the Matérn covariance function, which defines the decay rate of correlation with distance. The model is specified such that both spatial terms share the same range parameter.

### Model comparisons

For each taxonomic group and type of verticality (i.e., SES or mean verticality), we used restricted maximum likelihood (REML) to fit models with four different random effects structures [@zuurMixedEffectsModels2009]. The base model only included a spatial random field. Subsequent models additionally included biogeographic realm [sensu @dinersteinEcoregionBasedApproachProtecting2017] as a random intercept, a spatially varying coefficient for canopy height, and the combination of both. Including realm as a random intercept accounts for evolutionary differences among biogeographic realms that could lead to differences in verticality. The spatially varying coefficient allows the impact of canopy height on verticality to vary globally and allows us to test the hypothesis that canopy height has a positive effect on verticality only where climatic conditions support arboreality. We compared models using AIC and cross-validation. AIC is frequently used to identify the most parsimonious model [e.g., @barnettImprovingEstimatesSpecies2021]; however, the metric is not completely appropriate for hierarchical models (see Commander et al. 2022 `r word_comment(comment = "check citation", highlight = "Commander et al. 2022")`). Therefore, we additionally used cross-validation to evaluate out-of-sample predictive ability (*again check citation Commander et al. 2022*). We randomly assigned each data point to one of five folds, and then iteratively fit the model using four of the five folds while one fold was withheld for testing. The sum of the log-likelihoods for each left-out fold provides a measure of predictive ability that is compared across models. Cross validation was conducted using the ‘sdmTMB_cv’ function from the sdmTMB package. Models that included both a spatially varying coefficient for canopy height and geographic realm as a random intercept consistently had the lowest AIC value across all models, while model performance varied based on cross validation (Table S1 `r word_comment(comment = "add supp table", highlight = "Table S1")`). Given consistent performance based on AIC and our interest in the SVC, all predictions were made based on the model that included both realm and the SVC.

### Model validation

We validated models first using the ‘sanity’ function from the sdmTMB package. We additionally calculated quantile residuals using the DHARMa R package [@hartigDHARMaResidualDiagnostics2022] to visually assess that residuals were normally distributed and lacked patterning across space, rank transformed model predictions, raw model predictions.

To estimate uncertainty of SVCs for canopy height, we added draws of the SVC random field to draws of the fixed parameter for canopy height from 300 simulations. We then calculated the median and 0.025 and 0.975 quantiles to obtain a 95% confidence interval for each SVC estimate. Confidence intervals that did not overlap zero were considered significant [@andersonSdmTMBPackageFast2024].

@@

To examine whether the relationship between verticality and canopy height mediates the relationship between verticality and richness, we calculated Pearson correlations between mean or SES verticality and richness for each taxon considering all points, only points where canopy height significantly impacted verticality, and only points where canopy height had a positive significant effect on verticality.

@@

To examine potential impacts of climate change on assemblage verticality, we predicted mean and SES verticality under a future climate scenario using the models. However, we held vegetation variables constant at present-day values due to lack of global predictions of future canopy height and vegetation density. From the future predictions, we subtracted predictions to the data used to train the model (present predictions) and mapped the resulting differences. To standardize measures of change across taxa and between mean and SES verticality, we additionally calculated the percent relative change between future and present predictions for mean and SES verticality as $(verticality_{future,t,i} - verticality_{present,t,i})/(verticality_{max,t} - verticality_{min, t})*100$, where t indicates a given taxon and i indicates a given grid cell. Thus, negative values indicate predicted declines in verticality and positive values indicate predicted increases in verticality.

@@

### Climate

Spatial GLMMs indicated that verticality was limited by low precipitation across taxa, while impacts of high maximum temperatures, low minimum temperatures, and canopy height varied (Fig. \@ref(fig:mod-coefs)). Across taxa, precipitation of the driest and wettest months positively influenced mean and SES verticality (Fig. \@ref(fig:mod-coefs)). Minimum temperature of the coldest month positively impacted mammal and reptile verticality and did not significantly impact bird and amphibian verticality. Maximum temperatures of the warmest month exhibited a quadratic relationship with bird, reptile, and amphibian verticality and a positive linear relationship with mammals verticality. These patterns resulted in lower verticality at both high and low maximum temperatures in regions without high dry season precipitation for reptiles and amphibians, while verticality increased with temperature across a range of dry season precipitation for birds and mammals (Fig. S\@ref(fig:condeff-mean),\@ref(fig:condeff-ses)). Canopy height had globally positive impacts on SES verticality of reptiles and amphibians and mean verticality of amphibians, and vegetation density had no significant impact on mean or SES verticality of any taxon Fig. (\@ref(fig:mod-coefs)). However, impacts of canopy height varied spatially, exhibiting neutral impacts on verticality across much of the tropics and positive impacts in regions in regions with greater topographic relief, including the Andes, southern India, east-central Africa including the Eastern Arc Mountain Range, and Madagascar (Fig. \@ref(fig:svc), Fig. S\@ref(fig:svc-sig) ). Furthermore, environmental factors were unable to explain extremely high and low verticality as indicated by larger residuals when spatial random effects were excluded from model predictions (*supp figure*).

@@

Impacts of precipitation were particularly pronounced for amphibian verticality. In tropical wet forests, evapotranspiration allows amphibian body temperatures to remain below critical thermal maxima when exposed to high temperatures because water is consistently available [@sundayThermalsafetyMarginsNecessity2014]. However, in regions that experience pronounced dry seasons, amphibians behaviorally thermoregulate to avoid desiccation by increasing their use of understory habitats, which reduces assemblage verticality [@bashamVerticalStratificationCollapses2020]. Lower precipitation poses a similar constraint to arboreality for amphibians in dry regions at higher latitudes where basking to increase body temperatures can quickly lead to desiccation without adequate moisture availability or physiological defenses [@sundayThermalsafetyMarginsNecessity2014; @tracyNotJustSmall2010].

For mammals and reptiles, our models indicated that minimum cold temperatures had the greatest impact on verticality. In boreal forests, cold and windy conditions in the canopy pose large energetic costs to endotherms and can cause body temperatures to drop below critical thermal minima in ectotherms [@bymanEnergeticCostsWinter1988; @sundayThermalsafetyMarginsNecessity2014]. Species are therefore largely constrained to terrestrial or fossorial habitats, which are buffered from minimum temperature extremes [@defrenneGlobalBufferingTemperatures2019]. Yet despite the limiting or facilitative impacts that climate can have on verticality, climate conditions in combination with vegetation structure failed to predict extremely high and low verticality when the spatial random field was excluded from predictions, suggesting other factors contribute to unexplained variation.

@@

Our models indicated that canopy height has a variable and often insignificant impact on verticality, even across much of the tropics where tall, multi-layered forests would be expected to promote vertical expansion and high diversity of arboreal species [@burnsGriddedGEDIVegetation2024; @fengForestCanopyHeight2020; @rollLinkingVertebrateSpecies2015]. These results align with Basham et al. [-@bashamVerticalStratificationPatterns2023a], who similarly found an insignificant impact of canopy height on verticality within the tropics. Interestingly, previously reported spatial variation in the impact of canopy height on taxonomic and functional richness mirror, to some extent, spatial variation in the impacts of canopy height on verticality [@rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020; @coopsDisentanglingVegetationClimate2018]. For example, canopy height positively impacts verticality and species richness of amphibians in Madagascar, the Andes, Central America, Southern India, and New Guinea [@rollLinkingVertebrateSpecies2015]. Spatially concurrent positive impacts for birds occur in the Himalayas, New Guinea, and southeastern Australia, while for mammals they occur in northwestern Mexico and eastern Madagascar [@coopsDisentanglingVegetationClimate2018; @rollLinkingVertebrateSpecies2015]. Additionally, concurrent negative impacts for mammals occur in parts of the Amazon basin (\@ref(fig:svc-sig)) [@rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020]. Given spatial similarities in relationships between canopy height and both verticality and richness, we expected stronger positive correlations between verticality and richness in regions where canopy height positively impacted verticality. Yet, our results did not consistently support this hypothesis.

>>>>

---
author: ""
title: "Global patterns and drivers of vertebrate verticality"
bibliography: "`r rbbt::bbt_write_bib('docs/r-refs.bib', overwrite = TRUE)`"
output: 
  officedown::rdocx_document:
    reference_docx: "style-guide.docx"
    mapstyles:
      Normal: ['First Paragraph']
plots:
  caption:
    style: Image Caption
    pre: 'Figure '
    sep: ': '
---

@@

Vertical stacking of species is an emblematic signature of biologically rich regions and is associated with patterns of species richness from local to global scales [@macarthurPopulationEcologyWarblers1958; @macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019; @bashamVerticalStratificationPatterns2023a]. At local scales, species assemblages may be stratified across vertical resource and climate gradients from beneath the ground to the top of the canopy [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023]. Tall, multi-layered forests increase both habitat volume and niche opportunities within the habitat, acting effectively as multi-floored apartments that enable species to vertically partition resources [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]. By reducing competition, vertical stratification may be a mechanistic link to positive relationships between canopy height and species richness [@culbertInfluenceVerticalHorizontal2013; @fengForestCanopyHeight2020; @gillespieDistributionBirdSpecies2001; @oliveiraVerticalStratificationInfluences2019; @wagnerRemotelySensedTree2023]. However, canopy height does not consistently impact species richness at regional to global scales, with notable declines in the strength of the relationship at higher latitudes where the diversity of arboreal species generally declines [@larueDiversityVolumeRelationships2023; @rollLinkingVertebrateSpecies2015; @oliveiraVerticalStratificationInfluences2019]. Examining the impacts of environmental drivers of vertical stratification may thus add nuance to our understanding of the impact of canopy height on species richness.

At regional to global scales, climate may constrain vertical stratification in some regions despite the presence of tall forests. Vertical climate gradients are characterized by fossorial and terrestrial habitats that exhibit lower thermal variability and greater moisture availability than arboreal habitats [@scheffersTropicalMountainPasses2018; @leahyVerticalNicheElevation2021; @defrenneGlobalBufferingTemperatures2019]. At high latitudes, low minimum temperatures in the canopy may exceed physiological thresholds and constrain species to lower vertical positions [@oliveiraVerticalStratificationInfluences2019; @bakenSalamanderArborealityLimited2021]. In wet tropical forests, physiological constraints decline, which may allow species to stratify across the entire vertical habitat gradient and contribute to the high diversity of arboreal species found in these habitats [@oliveiraVerticalStratificationInfluences2019]. We therefore expect the impact of canopy height on verticality to vary across large spatial extents, exhibiting a positive impact where climate supports arboreal life strategies and a neutral impact where climatic conditions pose greater physiological challenges.

Spatial variation may also occur in the relationship between species richness and  verticality [@oliveiraVerticalStratificationInfluences2019]. Though prior studies have shown positive relationships between verticality and richness, the strength of this relationship is variable and generally more positive in tropical regions [@cabralFunctionalTraitsPhylogeny2022; @oliveiraVerticalStratificationInfluences2019]. Where climatic conditions allow canopy height to positively impact verticality, as may be expected in the wet tropics, the positive relationship between verticality and richness may be stronger because the addition of a vegetation layer allows the addition of arboreal species to the assemblage. In contrast, where taller canopies do not facilitate higher verticality, the relationship between verticality and richness may be weak.

As the climate changes, constraints on arboreal life strategies may shift with potential impacts on assemblage verticality and its association with species richness. In tropical lowlands, increasing temperatures and declining dry season precipitation (*cite*) may pose physiological challenges for arboreal animals and cause them to disperse toward higher elevations, lower vertical positions, or become locally extirpated [@scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019]. These scenarios would reduce assemblage verticality in the lowlands, reflecting climate-induced shifts in assemblage thermal tolerance [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024]. At high latitudes, increasing minimum temperatures could reduce climatic constraints on arboreal species and facilitate range shifts that increase verticality.

@@

Despite widespread agreement that verticality influences biodiversity patterns, global patterns of verticality, its environmental drivers, its relationship with species richness, and its vulnerability to climate change have not been comprehensively described for vertebrates at a global scale. To improve our understanding of the nuances of the relationships between climate, vegetation structure, verticality, and species richness, we integrate range maps of `r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians with climate and vegetation data to examine global patterns and environmental drivers of vertebrate verticality in wooded biomes. We then examine whether the relationship between verticality and canopy height mediates the relationship between verticality and species richness. Finally, we predict how climate change may impact global patterns of verticality.

@@

We mapped verticality and richness of vertebrate assemblages in wooded habitats [@dinersteinEcoregionBasedApproachProtecting2017] using 50 km grid cells and a cylindrical equal area projection (Supplementary methods). Grid cells with fewer than five species were excluded. Range maps were obtained from the IUCN red list for mammals and amphibians [@iucnIUCNRedList2023], BirdLife International for birds [@birdlifeinternationalandhandbookofthebirdsoftheworldBirdSpeciesDistribution2022], and the Global Assessment of Reptile Distributions (GARD) for reptiles [@rollGlobalDistributionTetrapods2017; @caetanoAutomatedAssessmentReveals2022]. For birds, we used breeding plus resident ranges. To calculate the verticality of each assemblage (i.e., grid cell), we assigned each species a verticality score between zero and one and took the average verticality of all species whose ranges overlapped with the grid cell. For mammals, reptiles, and amphibians, verticality scores were based on their use of fossorial (0), terrestrial (0.5), and arboreal (1) habitats as defined in the TetrapodTraits database [@mouraPhylogenyinformedCharacterisationGlobal2024]. For birds, we calculated verticality scores based on their use of different foraging strata as identified in the EltonTraits database [@wilmanEltonTraitsSpecieslevelForaging2014]. We classified foraging below the water surface, around the water surface, and on the ground as 0.5, in the understory as 0.667, mid-high canopy as 0.8337, and in or above the upper canopy as 1. If species used multiple vertical habitats, vertical scores were averaged. For example, if a species occurred in terrestrial (0.5) and arboreal (1) habitats, it would receive a verticality score of 0.75. Species that occupied only aerial vertical niches or that did not occur in terrestrial habitats were excluded from the analysis.

To examine spatial patterns of verticality independently of species richness, we calculated the standard effect size (SES) of mean verticality ($SES = (x_i - \bar{x})/sd(x)$, where x is a vector of expected mean verticality scores given random selection of species from a species pool). We defined species pools as species within all wooded habitats in each biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017] and generated a null distribution of expected verticality given random chance for each grid cell using 100 simulations.

@@

We obtained present and future global climate data from CHELSA v2.1 for mean daily maximum air temperature of the warmest month, mean daily minimum air temperature of the coldest month, temperature seasonality, precipitation of the wettest month, precipitation of the driest month, and precipitation seasonality [@kargerDataClimatologiesHigh2021; @kargerClimatologiesHighResolution2017]. Present climate data represents the time period 1981-2010 and future climate data represents an ensemble of climate scenarios for SSP585 for the time period 2071-2100. Vegetation data included canopy height [@langHighresolutionCanopyHeight2023], vegetation density [@crowtherMappingTreeDensity2015], and vegetation complexity (the product of canopy height and vegetation density). Additionally, we included historical climate velocity, which represents the average rate at which thermal isotherms have shifted between the last glacial maximum and present climate conditions [@sandelInfluenceLateQuaternary2011] and has been shown to influence global patterns of verticality [@oliveiraVerticalStratificationInfluences2019]. All environmental data were resampled to the resolution and projection of the species data.

@@

where $y_s$ is mean or SES verticality at location $s$ defined by x and y coordinates, $𝜇_s$ is the expected mean or SES verticality, and $𝑓^{-1}$ represents an inverse link function. We modeled SES verticality as a normal distribution and mean verticality as a beta distribution with a logit link function because values are constrained between 0 and 1. $𝜲^{main}_s$ is a vector of main effect covariates and 𝛽 represents a corresponding vector of estimated main effect coefficients. $α_g$ represents a random intercept defined by biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017], which accounts for potential variation in verticality drive by global evolutionary patterns. $⍵_s$ represents a spatial random field at location $s$, and describes additional spatial variation in the data not explained by the environmental covariates or random intercept. $𝜲^{svc}_sζ_s$ represents a spatially varying coefficient, where $𝜲^{svc}$ represents the vector of spatially varying parameter values and $ζ_s$ represents the spatially varying coefficients, which are described using a random field such that the coefficients follow a multivariate normal distribution with mean zero and covariance ($𝚺_ζ$). Both the spatial random field and the spatially varying coefficients are modeled as Gaussian random fields with a covariance matrix modeled by the Matérn covariance function, which defines the decay rate of correlation with distance. Gaussian random fields were approximated for models of mean and SES verticality for each taxa using a spatial mesh optimized such that the edge length was as large as possible to reduce computational time while remaining less than five times the spatial range of the model [@andersonSdmTMBPackageFast2024; @krainskiAdvancedSpatialModeling2019] (Supplementary methods). The model is specified such that both spatial terms share the same range parameter.

### Model comparison and evaluation

For each taxonomic group and type of verticality (i.e., SES or mean verticality), we used restricted maximum likelihood (REML) to fit models with four different random effects structures [@zuurMixedEffectsModels2009]. The base model only included a spatial random field. Subsequent models additionally included biogeographic realm [sensu @dinersteinEcoregionBasedApproachProtecting2017] as a random intercept, a spatially varying coefficient for canopy height, and the combination of both. Including realm as a random intercept accounts for evolutionary differences among biogeographic realms that could lead to differences in verticality. The spatially varying coefficient allows the impact of canopy height on verticality to vary globally and allows us to test the hypothesis that canopy height has a positive effect on verticality only where climatic conditions support arboreality. We compared models using AIC and cross-validation. AIC is frequently used to identify the most parsimonious model [e.g., @barnettImprovingEstimatesSpecies2021]; however, the metric is not completely appropriate for hierarchical models (see Commander et al. 2022 `r word_comment(comment = "check citation", highlight = "Commander et al. 2022")`). Therefore, we additionally used cross-validation to evaluate out-of-sample predictive ability (*again check citation Commander et al. 2022*). We randomly assigned each data point to one of five folds, and then iteratively fit the model using four of the five folds while one fold was withheld for testing. The sum of the log-likelihoods for each left-out fold provides a measure of predictive ability that is compared across models. Cross validation was conducted using the ‘sdmTMB_cv’ function from the sdmTMB package. Models that included both a spatially varying coefficient for canopy height and geographic realm as a random intercept consistently had the lowest AIC value across all models, while model performance varied based on cross validation (Tabe S\@ref(tab:mod-compare-tab)). Given consistent performance based on AIC and our interest in the SVC, all predictions were made based on the model that included both realm and the SVC.

We validated models first using the ‘sanity’ function from the sdmTMB package. We additionally calculated quantile residuals using the DHARMa R package [@hartigDHARMaResidualDiagnostics2022] to visually assess that residuals were normally distributed and lacked patterning across space and to examine any patterns across rank transformed and raw model predictions.

We estimated uncertainty of SVCs for canopy height by adding draws of the SVC random field to draws of the fixed parameter for canopy height from 300 simulations. We then calculated the median and 0.025 and 0.975 quantiles to obtain a 95% confidence interval for each SVC estimate. Confidence intervals that did not overlap zero were considered significant [@andersonSdmTMBPackageFast2024].

@@

To examine whether the relationship between verticality and canopy height mediates the relationship between verticality and richness, we calculated Pearson correlations between mean or SES verticality and richness for each taxon considering all points and only points where canopy height had a positive significant effect on verticality.

@@

To examine potential impacts of climate change on assemblage verticality, we used the models to predict mean and SES verticality under a future climate scenario. We held vegetation variables constant at present-day values due to lack of global predictions of future canopy height and vegetation density. We subtracted predictions to the present-day climate data used to train the model from future predictions and mapped the resulting differences. To standardize measures of change across taxa and between mean and SES verticality, we calculated the percent relative change between present and future predictions for mean and SES verticality as $(verticality_{future,t,i} - verticality_{present,t,i})/(verticality_{max,t} - verticality_{min, t})*100$, where t indicates a given taxon and i indicates a given grid cell. Negative values indicate predicted declines in verticality and positive values indicate predicted increases in verticality.

@@

Spatial GLMMs indicated that verticality was limited by low precipitation across taxa, while impacts of high maximum temperatures, low minimum temperatures, and canopy height varied (Fig. \@ref(fig:mod-coefs)). Across taxa, precipitation of the driest and wettest months positively influenced mean and SES verticality (Fig. \@ref(fig:mod-coefs)). Minimum temperature of the coldest month positively impacted mammal and reptile verticality and did not significantly impact bird and amphibian verticality. Maximum temperatures of the warmest month exhibited a quadratic relationship with bird, reptile, and amphibian verticality and a positive linear relationship with mammal verticality. These patterns resulted in lower verticality at both high and low maximum temperatures in regions without high dry season precipitation for reptiles and amphibians, while verticality increased with temperature across a range of dry season precipitation for birds and mammals (Fig. S\@ref(fig:condeff-mean),\@ref(fig:condeff-ses)). Canopy height had globally positive impacts on SES verticality of reptiles and amphibians and mean verticality of amphibians, and vegetation density had no significant impact on mean or SES verticality of any taxon Fig. (\@ref(fig:mod-coefs)). However, impacts of canopy height varied spatially, exhibiting neutral impacts on verticality across much of the tropics and positive impacts in regions in regions with greater topographic relief, including the Andes, southern India, east-central Africa including the Eastern Arc Mountain Range, and Madagascar (Fig. \@ref(fig:svc), Fig. S\@ref(fig:svc-sig) ). Furthermore, environmental factors were unable to explain extremely high and low verticality as indicated by larger residuals (Fig. S\@ref(fig:resid-plt)).

@@

Impacts of precipitation were particularly pronounced for amphibian verticality. In tropical wet forests, consistent water availability allows amphibians to maintain body temperatures below critical thermal maxima via evapotranspiration [@sundayThermalsafetyMarginsNecessity2014]. However, in regions that experience pronounced dry seasons, amphibians may behaviorally thermoregulate to avoid desiccation by increasing their use of understory habitats, which reduces assemblage verticality [@bashamVerticalStratificationCollapses2020]. Lower precipitation poses a similar constraint to arboreality for amphibians in dry regions at higher latitudes where basking to increase body temperatures can quickly lead to desiccation without adequate moisture availability or physiological defenses [@sundayThermalsafetyMarginsNecessity2014; @tracyNotJustSmall2010].

For mammals and reptiles, our models indicated that minimum cold temperatures had the greatest impact on verticality. In boreal forests, cold and windy conditions in the canopy pose large energetic costs to endotherms and can cause body temperatures to drop below critical thermal minima in ectotherms [@bymanEnergeticCostsWinter1988; @sundayThermalsafetyMarginsNecessity2014]. Species are therefore largely constrained to terrestrial or fossorial habitats, which are buffered from minimum temperature extremes [@defrenneGlobalBufferingTemperatures2019]. Yet despite the limiting or facilitative impacts that climate can have on verticality, climate conditions in combination with vegetation structure failed to predict extremely high and low verticality when the spatial random field was excluded from predictions, suggesting other factors influence global patterns of verticality.

@@

Our models indicated that canopy height has a variable and often insignificant impact on verticality, even across much of the tropics where tall, multi-layered forests would be expected to promote vertical expansion and high diversity of arboreal species [@burnsGriddedGEDIVegetation2024; @fengForestCanopyHeight2020; @rollLinkingVertebrateSpecies2015]. These results correspond to Basham et al. [-@bashamVerticalStratificationPatterns2023a], who similarly found an insignificant impact of canopy height on verticality within the tropics. Interestingly, previously reported spatial variation in the impact of canopy height on taxonomic and functional richness mirror, to some extent, spatial variation in the impacts of canopy height on verticality [@rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020; @coopsDisentanglingVegetationClimate2018]. For example, canopy height positively impacts verticality and species richness of amphibians in Madagascar, the Andes, Central America, Southern India, and New Guinea [@rollLinkingVertebrateSpecies2015]. Spatially concurrent positive impacts for birds occur in the Himalayas, New Guinea, and southeastern Australia, while for mammals they occur in northwestern Mexico and eastern Madagascar [@coopsDisentanglingVegetationClimate2018; @rollLinkingVertebrateSpecies2015]. Additionally, concurrent negative impacts for mammals occur in parts of the Amazon basin (Fig. S\@ref(fig:svc-sig)) [@rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020]. Given spatial similarities in relationships between canopy height and both verticality and richness, we expected stronger positive correlations between verticality and richness in regions where canopy height positively impacted verticality. Yet, our results did not consistently support this hypothesis.
